Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in \textit{Erythranthe cardinalis}

### 2017-08-30 ms/ms.Rnw
minor change
restructuring

  # The path to the ms directory may need to be changed for different users
  # Manuscript files
  pathMS <- "~/Google Drive/CardAdapt/ms" # Chris' MacBook Air
  # pathMS <- "~/Documents/CardAdapt/ms" # Chris' Linux laptop
  # pathMS <- "/Users/amylauren/Desktop/CardAdapt/ms" # Amy's desktop

@@

% Intrinsic versus plastic variation 
When populations are locally adapted, reaction norms for fitness will cross, such that local genotypes have higher fitness than foreign genotypes and rank orders change across environments \citep{Kawecki_Ebert_2004}. The traits that underlie local adaptation, however, need not mirror this pattern. Populations can have fixed genetic differences conferring trait values that are adaptive at home but neutral or maladaptive away. Alternatively, the ability to plastically respond to a particular environment or the magnitude of response to an environment could be adaptive. We distinguish between these patterns of adaptive trait differences by referring to `intrinsic' and `plastic' trait variation, respectively. Both intrinsic and plastic trait variation can be explained by genetic differences and both are involved in adaptation. For example, intrinsic differences in photoperiod responses \citep{Blackman_etal_2011} and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment. Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype-by-environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}. In plants especially, we know little about the prevalence and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance.

@@

We imposed four treatments, a fully-factorial cross of two temperature levels and two watering levels. The temperature levels closely simulated an average growing season at the thermal extremes of the species range, which we designate as Hot and Cool treatments. Watering levels contrasted a perennial and seasonal stream, which we refer to as Well-watered and Drought treatments. A detailed description of treatments is provided in the Supplemental Materials and Methods and summarized in Fig~\ref{fig:Fig_ExptlDes}. Because growth chambers cannot be subdivided, one chamber was assigned to the Hot treatment level and another to the Cool treatment level. Within each chamber, there were two Well-watered blocks and two Drought blocks. The photosynthetically active radiation in both chambers was approximately 400 $\mu$mol quanta m$^{-2}$ s$^{-1}$. The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature levels (data not shown). Lower humidity would have made the drought more severe, but low soil moisture is stressful in and of itself. The total number of plants in each treatment was: $n_{\text{cool,dry}} = \Sexpr{sum(indPerPop["Cool", "Dry", ])}$; $n_{\text{cool,ww}} = \Sexpr{sum(indPerPop["Cool", "Wet", ])}$; $n_{\text{hot,dry}} = \Sexpr{sum(indPerPop["Hot", "Dry", ])}$; $n_{\text{hot,ww}} = \Sexpr{sum(indPerPop["Hot", "Wet", ])}$. Each population had \Sexpr{min(indPerPop)}--\Sexpr{max(indPerPop)} individuals per treatment level (mean = \Sexpr{round(mean(indPerPop), 1)}, median = \Sexpr{median(indPerPop)}).

@@

  pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/DroughtResponse.pdf", w = 5.5, h = 5, useDingbats = F)
  par(mai = c(1, 1.5, 0.25, 0.25))

  plot(focClim$bio16_avg, x[, "Dry", "Hot"] / x[, "Wet", "Hot"], axes = F, frame.plot = T,
       pch = 21, col = "black", bg = primShade0, cex = 2, type = "n",
       xlab = "Precipitation of Wettest Quarter (mm)", 
       ylab = "Relative Survivorship\n(Hot/Dry : Hot/Wet)", cex.lab = 1.5)
  points(focClim$bio16_avg, x[, "Dry", "Hot"] / x[, "Wet", "Hot"], pch = 21, col = "black", 
         bg = primShade0, cex = 1.5, lwd = 2)
  axis(1, lwd = 0, lwd.ticks = 1)
  axis(2, lwd = 0, lwd.ticks = 1, las = 1)
  dev.off()

@@

  ##### TEMPORARY VERSION FOR TALK #####
  # Plot latitude versus PC1
  pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/Figure_PC1vLat.pdf", 5.5, 5, useDingbats = F)
  par(mai = c(1, 1.5, 0.25, 0.25))
  plot(focClim$Lat, focClim$PC1, xlim = c(32, 44), ylim = c(-4, 4), axes = F,
    frame.plot = T, type = "n", xlab = "", ylab = "")
  title(xlab = "Latitude", cex.lab = 1.5)
  title(ylab = "TraitPC1", cex.lab = 2, line = 5)
  title(ylab = substitute(plain(Principal~Component~1)~group("(", x, ")"), 
    list(x = pve)))
  axis(1, at = seq(32, 44, 4), labels = c(expression(32*degree), 
		expression(36*degree), expression(40*degree), expression(44*degree)), 
		lwd = 0, lwd.ticks = 1) 
  axis(2, at = seq(-4, 4, 2), las = 1, lwd = 0, lwd.ticks = 1)    
  fitLat <- lm(PC1 ~ Latitude, data = focClim)

@@

@@

<<Fig_VarSelect, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=

@@

##### TEMPORARRY FIGURE FOR TALK #####

path.root <- "~/Google Drive/CardAdapt" 
setwd(path.root)
source("Data/Climate/Rcode/1_Setup_directories.R")

load(paste(path.obj, "/AllOcc_avg1981-2010.RData", sep = "")) # all.avg30
load(paste(path.obj, "/AllOcc_cva1981-2010.RData", sep = "")) # all.cva30
load(paste(path.obj, "/AllOccSA_avg1981-2010_62km.RData", sep = "")) # allSA.avg
load(paste(path.obj, "/AllOccSA_cva1981-2010_62km.RData", sep = "")) # allSA.cva

# Change column names for SA variables
sharedCols <- c("Latitude", "Longitude", "Elevation", "PopID", "ID", "PopID")
colnames(allSA.avg)[!colnames(allSA.avg) %in% sharedCols] <-
  paste0(colnames(allSA.avg)[!colnames(allSA.avg) %in% sharedCols], "_SA")

colnames(allSA.cva)[!colnames(allSA.cva) %in% sharedCols] <-
  paste0(colnames(allSA.cva)[!colnames(allSA.cva) %in% sharedCols], "_SA")

# Combine into single data.frame
allClim <- cbind(all.avg30, all.cva30[rownames(all.avg30), !(colnames(all.cva30) %in% sharedCols)],
                 allSA.avg[rownames(all.avg30), !(colnames(allSA.avg) %in% sharedCols)],
                 allSA.cva[rownames(all.avg30), !(colnames(allSA.cva) %in% sharedCols)])

  pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/ClimateSpace.pdf", w = 5.5, h = 5, useDingbats = F)
  par(mai = c(1, 1.5, 0.25, 0.25))

  with(allClim, plot(bio10_avg - 273.16, bio16_avg, xlim = c(15, 27), ylim = c(100, 1500),
                     xlab = "", ylab = "", axes = F, frame.plot = T, col = primShade1, pch = 19))
  with(focClim, points(bio10_avg - 273.16, bio16_avg, cex = 2, col = "black", bg = primShade0, 
                       lwd = 2, pch = 21))
  axis(1, at = seq(16, 26, 2), labels = c(expression(16*degree), expression(18*degree), 
                                          expression(20*degree), expression(22*degree), 
                                          expression(24*degree), expression(26*degree)),
		lwd = 0, lwd.ticks = 1) 
  axis(2, las = 1, lwd = 0, lwd.ticks = 1)
  title(xlab = "Temperature of Warmest Quarter", cex.lab = 1.5, line = 2.5)
  title(xlab = expression(paste("(", degree, "C, 1981-2010 average)")), cex.lab = 1, line = 4)
  title(ylab = "Precipitation of Wettest Quarter", line = 5, cex.lab = 1.5)
  title(ylab = "(mm, 1981-2010 average)", cex.lab = 1, line = 4)

  legend(grconvertX(0.96, "nfc", "user"), grconvertY(0.95, "nfc", "user"), xjust = 1, 
         legend = c("all recorded populations", "focal populations"),
         pch = c(19, 21), col = c(primShade1, "black"), pt.bg = primShade0, pt.cex = c(1, 2))
dev.off()

##### TEMPORARY FOR REVIEW RESPONSE #####
  # This predicts that the difference between local and neighbourhood climate estimates should be lower 
# for northern populations than all others

# Separate into different datasets
c1 <- !grepl("_SA", climVars)
c2 <- grepl("_SA", climVars)

dat1 <- thinnedOccClimate[, climVars[c1]]
dat2 <- thinnedOccClimate[, climVars[c2]]

# Standardize
for (i in ncol(dat1)) {
  x <- c(dat1[, i], dat2[, i])
  x <- (x - mean(x)) / sd(x)
  dat1[, i] <- x[1:nrow(dat1)]
  dat2[, i] <- x[(nrow(dat1) + 1):(2 * nrow(dat1))]
}

# Calculate difference
d12 <- dat1
r <- numeric(ncol(d12))
for (i in 1:ncol(d12)) {
  d12[, i] <- abs(dat2[, i] - dat1[, i])
  r[i] <- cor(thinnedOccClimate$Latitude, d12[, i])
}
hist(r[1:24])
hist(r[25:48])
summary(r)

plot(thinnedOccClimate$Latitude, apply(d12, 1, function(X) sum(abs(X))))
plot(thinnedOccClimate$Latitude, d12$bio12_avg)
plot(thinnedOccClimate$Latitude, d12$bio12_cva)
plot(thinnedOccClimate$Latitude, d12$bio1_avg)
plot(thinnedOccClimate$Latitude, d12$bio1_cva)

@@

\bibliography{../../CommonResources/refs}
\bibliographystyle{evolution}

>>>>

  # The path to the ms directory may need to be changed for different users
  # Manuscript files
  pathMS <- "~/Google Drive/CardAdapt/ms" # Chris' MacBook Air
  # pathMS <- "~/Documents/Temporary/ms" # Chris' Poisson laptop
  # pathMS <- "/Users/amylauren/Desktop/CardAdapt/ms" # Amy's desktop

@@

% Intrinsic versus plastic variation 
When populations are locally adapted, reaction norms for fitness will cross, such that local genotypes have higher fitness than foreign genotypes and rank orders change across environments \citep{Kawecki_Ebert_2004}. The traits that underlie local adaptation, however, need not mirror this pattern. Populations can have fixed genetic differences conferring trait values that are adaptive at home but neutral or maladaptive away. Alternatively, the ability to plastically respond to a particular environment or the magnitude of response to an environment could be adaptive. We distinguish between these patterns of adaptive trait differences by referring to `intrinsic' and `plastic' trait variation, respectively. Both intrinsic and plastic trait variation can be explained by genetic differences and both are involved in adaptation. For example, intrinsic differences in photoperiod responses \citep{Blackman_etal_2011} and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment. Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype-by-environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}. In plants especially, we know little about the prevalence and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance \citep{Flood_etal_2011}.

@@

We imposed four treatments, a fully-factorial cross of two temperature levels and two watering levels. The temperature levels closely simulated an average growing season at the thermal extremes of the species range, which we designate as Hot and Cool treatments. Watering levels contrasted a perennial and seasonal stream, which we refer to as Well-watered and Drought treatments. A detailed description of treatments is provided in the Supplemental Materials and Methods and summarized in Fig~\ref{fig:Fig_ExptlDes}. Because growth chambers cannot be subdivided, one chamber was assigned to the Hot treatment level and another to the Cool treatment level. Within each chamber, there were two Well-watered blocks and two Drought blocks. The photosynthetically active radiation in both chambers was approximately 400 $\mu$mol quanta m$^{-2}$ s$^{-1}$ and set to a 16:8 light:dark cycle to simulate summer growing conditions. The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature levels (data not shown). Lower humidity would have made the drought more severe, but low soil moisture is stressful in and of itself. The total number of plants in each treatment was: $n_{\text{cool,dry}} = \Sexpr{sum(indPerPop["Cool", "Dry", ])}$; $n_{\text{cool,ww}} = \Sexpr{sum(indPerPop["Cool", "Wet", ])}$; $n_{\text{hot,dry}} = \Sexpr{sum(indPerPop["Hot", "Dry", ])}$; $n_{\text{hot,ww}} = \Sexpr{sum(indPerPop["Hot", "Wet", ])}$. Each population had \Sexpr{min(indPerPop)}--\Sexpr{max(indPerPop)} individuals per treatment level (mean = \Sexpr{round(mean(indPerPop), 1)}, median = \Sexpr{median(indPerPop)}).

@@

  #pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/DroughtResponse.pdf", w = 5.5, h = 5, useDingbats = F)
  par(mai = c(1, 1.5, 0.25, 0.25))

  plot(focClim$bio16_avg, x[, "Dry", "Hot"] / x[, "Wet", "Hot"], axes = F, frame.plot = T,
       pch = 21, col = "black", bg = primShade0, cex = 2, type = "n",
       xlab = "Precipitation of Wettest Quarter (mm)", 
       ylab = "Relative Survivorship\n(Hot/Dry : Hot/Wet)", cex.lab = 1.5)
  points(focClim$bio16_avg, x[, "Dry", "Hot"] / x[, "Wet", "Hot"], pch = 21, col = "black", 
         bg = primShade0, cex = 1.5, lwd = 2)
  axis(1, lwd = 0, lwd.ticks = 1)
  axis(2, lwd = 0, lwd.ticks = 1, las = 1)
  #dev.off()

@@

  ##### TEMPORARY VERSION FOR TALK #####
  # Plot latitude versus PC1
  #pdf("~/Google Drive/Talks/JobTalk/3_Mimulus/Figure_PC1vLat.pdf", 5.5, 5, useDingbats = F)
  par(mai = c(1, 1.5, 0.25, 0.25))
  plot(focClim$Lat, focClim$PC1, xlim = c(32, 44), ylim = c(-4, 4), axes = F,
    frame.plot = T, type = "n", xlab = "", ylab = "")
  title(xlab = "Latitude", cex.lab = 1.5)
  title(ylab = "TraitPC1", cex.lab = 2, line = 5)
  title(ylab = substitute(plain(Principal~Component~1)~group("(", x, ")"), 
    list(x = pve)))
  axis(1, at = seq(32, 44, 4), labels = c(expression(32*degree), 
		expression(36*degree), expression(40*degree), expression(44*degree)), 
		lwd = 0, lwd.ticks = 1) 
  axis(2, at = seq(-4, 4, 2), las = 1, lwd = 0, lwd.ticks = 1)    
  fitLat <- lm(PC1 ~ Latitude, data = focClim)

@@

 # dev.off()

@@

<<Fig_VarSelect, eval=FALSE, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=

@@

@@

\bibliography{refs}
\bibliographystyle{evolution}

### 2017-08-30 ms/ms.tex
restructuring

We used 16 populations from throughout the range of \textit{E. cardinalis} (Table~\ref{table:Table_FocalPops}). These populations were intentionally chosen to span much of the climatic range of the species based on all known occurrences (see below). Seeds were collected in the field from mature, undehisced fruit left open for 2-4 weeks to dry, then stored at room temperature. We used seeds from 154 families, 4--12 (mean = 9.6, median = 12) families per population.

@@

We imposed four treatments, a fully-factorial cross of two temperature levels and two watering levels. The temperature levels closely simulated an average growing season at the thermal extremes of the species range, which we designate as Hot and Cool treatments. Watering levels contrasted a perennial and seasonal stream, which we refer to as Well-watered and Drought treatments. A detailed description of treatments is provided in the Supplemental Materials and Methods and summarized in Fig~\ref{fig:Fig_ExptlDes}. Because growth chambers cannot be subdivided, one chamber was assigned to the Hot treatment level and another to the Cool treatment level. Within each chamber, there were two Well-watered blocks and two Drought blocks. The photosynthetically active radiation in both chambers was approximately 400 $\mu$mol quanta m$^{-2}$ s$^{-1}$. The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature levels (data not shown). Lower humidity would have made the drought more severe, but low soil moisture is stressful in and of itself. The total number of plants in each treatment was: $n_{\text{cool,dry}} = 169$; $n_{\text{cool,ww}} = 174$; $n_{\text{hot,dry}} = 176$; $n_{\text{hot,ww}} = 183$. Each population had 8--12 individuals per treatment level (mean = 11, median = 11).

>>>>

\textit{E. cardinalis} is a perennial forb native to the Western US (California and Oregon). It is predomintantly outcrossing, self-compatible, and pollinated primarily by hummingbirds. We used 16 populations from throughout the range of \textit{E. cardinalis} (Table~\ref{table:Table_FocalPops}). These populations were intentionally chosen to span much of the climatic range of the species based on all known occurrences (see below). Seeds were collected in the field from mature, undehisced fruit left open for 2-4 weeks to dry, then stored at room temperature. We used seeds from 154 families, 4--12 (mean = 9.6, median = 12) families per population.

@@

We imposed four treatments, a fully-factorial cross of two temperature levels and two watering levels. The temperature levels closely simulated an average growing season at the thermal extremes of the species range, which we designate as Hot and Cool treatments. Watering levels contrasted a perennial and seasonal stream, which we refer to as Well-watered and Drought treatments. A detailed description of treatments is provided in the Supplemental Materials and Methods and summarized in Fig~\ref{fig:Fig_ExptlDes}. Because growth chambers cannot be subdivided, one chamber was assigned to the Hot treatment level and another to the Cool treatment level. Within each chamber, there were two Well-watered blocks and two Drought blocks. The photosynthetically active radiation in both chambers was approximately 400 $\mu$mol quanta m$^{-2}$ s$^{-1}$ and set to a 16:8 light:dark cycle to simulate summer growing conditions. The growth chambers did not control humidity, but because of watering and high plant transpiration rates, the relative humidity was quite high in both temperature levels (data not shown). Lower humidity would have made the drought more severe, but low soil moisture is stressful in and of itself. The total number of plants in each treatment was: $n_{\text{cool,dry}} = 169$; $n_{\text{cool,ww}} = 174$; $n_{\text{hot,dry}} = 176$; $n_{\text{hot,ww}} = 183$. Each population had 8--12 individuals per treatment level (mean = 11, median = 11).
